\(Q\)-Bosons and their relations with integrable hierarchies

### 2024-04-16 draft.tex
add correlation functions

@@

\title[\(Q\)-Bosons and relations with integrable hierarchies]{Comments on \(Q\)-Bosons,
and their relations with integrable hierarchies}

@@

\address{\noindent 
Instituto de Física Teórica, UNESP-Universidade Estadual Paulista
R. Dr. Bento T. Ferraz 271, Bl. II, Sao Paulo 01140-070, SP, Brazil\\
and
Instituto de Física\\ Universidade de S\~ao Paulo\\ 
Rua do Matão Travessa 1371, 05508-090\\ São Paulo, SP. Brazil
}

@@

\begin{abstract}
In this work, we study aspects of the \(Q\)-boson model, and 
its relations with integrable hierarchies. 
\vspace{4pt}
\noindent \textbf{Keywords:} Integrability, Q-bosons, Schur

\keywords{integrability, spin-chains}
\subjclass[2020]{82B20, 82B23}
\date{\today}

\end{abstract}

\maketitle

\setcounter{tocdepth}{3}
\tableofcontents

@@

Bogoliubov~\cite{Bogoliubov2005}, has shown that the scalar product of
two vectors in the \(N\)-particles sector in a chain of length \(M+1\)
is
\begin{equation}
\begin{split}
\label{eq:scalar}
  S(N, M|\bm{x}, \bm{y}) & =
  \bra{\bm{0}} \prod_{j=1}^N \mathbb{C}(x_j) \prod_{j=1}^N \mathbb{B}(y_j) \ket{\bm{0}} \\ 
 & = \frac{ \det H(\bm{x},\bm{y})}{ \prod_{i<j}(x_i - x_j)(y_i - y_j)} \; ,
\end{split}
\end{equation}
where \(H\) is an \(N\times N\) matrix with components
\begin{equation}
  H_{ij} = H(x_i, y_j) 
  =\frac{1 - (x_i y_j)^{ M + N}}{1 - x_i y_j }\; .
\end{equation}

@@

\subsubsection{Dyson gases}

It is also interesting to consider the particular case when \(M\to
\infty\), but with a finite number of particles \(N\). The
partititions \(\lambda \in [N, \infty]\) satisfy the condition
\(\ell(\lambda) \leq N\), therefore, the scalar
product~(\ref{eq:scalar_exp}) becomes
\begin{equation}
  S_N(\bm{t}, \bm{t}') = \sum_{\substack{\lambda \\ \ell(\lambda) \leq N}}
  s_\lambda(\bm{t})  s_\lambda(\bm{t}') \; .
\end{equation}
From~\cite{Zabrodin:2010ii}, we know that this expression can be
written as the matrix integral
\begin{equation}
  S_N(\bm{t}, \bm{t}') =
  \frac{1}{N!} \prod_{\ell=1}^N \oint_{\Gamma_\ell} \frac{dz_\ell}{2 \pi i z_\ell}
  e^{\xi(\bm{t}, z_\ell) - \xi(\bm{t}', z_{\ell}^{-1})} \Delta(z)\Delta(z^{-1})\; ,
\end{equation}
where
\begin{equation}
\xi(\bm{t}, z) = \sum_{k\geq 0} t_k z^k \qquad
\xi(\bm{t}', 1/z) = \sum_{k\geq 0} t'_{-k} z^{-k} \; . 
\end{equation}
See also~\cite{Kharchev:1991gd} for a detailed proof of this relation,
and~\cite{Orlov:2005} for other details.

What is very interesting about this result is what it implies when we 
impose the Bethe equations~(\ref{eq:bethe_eq}). Suppose that 
\(x_j = e^{-ip_j} \in \mathbb{S}^1\), and we set \(\bm{t}' = - \bm{t}^\star\).
In this particular case, we have
\begin{equation}
\xi(\bm{t}, z) - \xi(\bm{t}', 1/z)  = 2 \ \textrm{Re}\left(\sum_k t_k z^k\right)\; .
\end{equation}
Then, the phase model is equivalent to an ensemble of \(N\)
2D Coulomb particles on a circle.

\subsubsection{KP tau function}
Finally, one may also observe that if we fix one set of coordinates,
say \(\bm{y}\), then we write
\begin{equation}
 S(N,M|\{\bm{t}\}) 
 = \sum_{\lambda} s_\lambda(\bm{y})  s_\lambda(\bm{t}) 
 \equiv \sum_{\lambda} c_\lambda(\bm{y})  s_\lambda(\bm{t}) \; ,
\end{equation}
with coefficients \(c_\lambda(\bm{y}) = \det (h_{\lambda_i-i +j}(\bm{y}))\).
But now, it is a trivial to notice that these are Plücker
coordinates in the Jacobi-Trudi form, see~\cite{Miwa2000, Alexandrov:2012tr}.

@@

\subsection{Correlation functions}
Bogoliubov has also shown that the correlation functions
\begin{subequations}
\begin{equation}
  P(m|\bm{x}, \bm{y}) = \bra{0} \prod_{j=1}^N C(x) \prod_{j=1}^{N-1} B(y) \phi_m^\dagger \ket{0}
\end{equation}
can be written as
\begin{equation}
  P(m|\bm{x}, \bm{y}) = \frac{(-1)^{N-1}}{(\sqrt{y_N}^{N-1}}
  \left( \prod_{t<N} \frac{x_N - y_t}{y_t} \right) \frac{\det Q}{\det H} S(N,M|\bm{x}, \bm{y})
\end{equation}
\end{subequations}
where the \(N\times N\) matrix \(Q\) has components 
\begin{equation}
 Q_{jN} = x_j^{(M + N - 1- 2m)/2} \quad  \textrm{and} \quad 
 Q_{jk} = H_{jk} \; , 
\end{equation}
where \(H_{jk} = H(x_j, y_k)\) are the components of the matrix \(H\). 
Since the components \(Q_{jN}\) are independent of the coordinates \(y\), 
we cannot write the above expression as a Toda hierarchy tau
function.

We already know that 
\begin{equation}
   S(N,M|\bm{x}, \bm{y}) =  \Delta(x) \Delta(y) \det H \; ,
\end{equation}
therefore we define
\begin{equation}
  \mathbb{P}(m|\bm{x}, \bm{y}) =\frac{\det Q}{\Delta(x)}\; ,
\end{equation}
and we are going to treat the coordinates \(\{ \bm{y} \}\) as
a set of \(N\) fixed constants, in a such a way that we write \(N\)
functions \(\bm{F}(z) = (F_1, \dots, F_N)\) defined as
\begin{equation}
  F_j (z) = H(z, y_j) \quad \textrm{if} \ j \neq N\; , \qquad \textrm{and}\qquad 
   F_N (z)  = z^{(M + N - 1 - 2m)/2} 
\end{equation}
Moreover, we have that
\begin{equation}
  H(z, y_j) = \frac{1 - (y_j z)^{M + N}}{1 - y_j z} = \sum_{n=0}^{N + M - 1} (y_j z)^n\; , 
\end{equation}
therefore, we write
\begin{equation}
  F_j(z) =  \sum_{n=0}^{M + N -1} f_{j, n} z^n
\end{equation}
where we will also assume the \(M + N \in 2 \mathbb{Z}\). With these definitions,
we conclude that
\begin{equation}
  \mathbb{P}(m|\bm{x}, \bm{y}) =\frac{\det_{jk} F_j(x_k)}{\Delta(x)}\; ,
\end{equation}
and we write~\cite{Takasaki:2010qm}, 
\begin{equation}
  \det_{jk} F_j(x_k) = \det_{jk} \left(  \sum_{n=0}^{M + N -1} f_{j, n} x_k^n \right) \; .
\end{equation}
With the aid of the Cauchy-Binet formula, we write  
\begin{equation}
  \mathbb{P}(m|\bm{x}, \bm{y}) = \sum_{0\leq \ell_N\leq \dots \leq \ell_1 \leq N+M}
  \frac{\det_{jk}(f_{j, \ell_k}) \det_{jk}(x_k^{\ell_j})}{\Delta(x)}\; ,
\end{equation}
and from the definition of the Schur polynomials and the
Jacobi-Trudi expression of Plücker coordinates, we can write this last
expression as 
\begin{equation}
  \mathbb{P}(m|\bm{x}, \bm{y}) = \sum_{\lambda} c_\lambda(\bm{y}) s_\lambda(\bm{x}) \; ,
\end{equation}
where \(c_\lambda(f_{j, \ell_k})\),  \(\ell_k = \lambda_k - k +N\), define point
in the finite dimensional Grassmannian manifold \(\mathrm{Gr}(N, N+M)\).

\subsubsection{Not everything is a tau-function}
It is also worth noticing that not everything is a tau function. Consider the state
\begin{equation}
  \ket{\mathcal{Y}} =
 \sum_{\vec{n}} \prod_{j=0}^M \ket{n_j}\quad \sum_j n_j = N\; .
\end{equation}
We have~\cite{Bogoliubov2005} the correlation function 
\begin{equation}
  \prod_{j=1}^N x^{M/2} \bra{\mathcal{Y}} \prod_{j=1}^N B(\bm{x}) \ket{\bm{0}} = 
  \sum_{\lambda\subseteq (N,N,M)} s_\lambda(\bm{x})\; . 
\end{equation}
This sum of Schur polynomials are not tau functions of the KP hierarchy.

>>>>

\usepackage{amsaddr} % affiliation in the first page

@@

\title[\(Q\)-Bosons and relations with integrable hierarchies]{\(Q\)-Bosons
and their relations with integrable hierarchies}

@@

\address{\noindent 
Instituto de Física Teórica, UNESP-Universidade Estadual Paulista,
R. Dr. Bento T. Ferraz 271, Bl. II, Sao Paulo 01140-070, SP, Brazil\\
\&
Instituto de Física, Universidade de S\~ao Paulo,
Rua do Matão Travessa 1371, 05508-090 São Paulo, SP. Brazil
}

@@

\begin{abstract}
In this work, we study aspects of the \(Q\)-boson model, and 
its relations with integrable hierarchies. 
\end{abstract}

%\date{\today}
\keywords{Integrability, Bethe, Schur}
\subjclass[2020]{82B20, 82B23}
\maketitle

\setcounter{tocdepth}{2}
\tableofcontents

@@

Bogoliubov~\cite{Bogoliubov2005}, has shown that the scalar product of
two vectors in the \(N\)-particles sector in a chain of length \(M+1\)
is
\begin{equation}
\begin{split}
\label{eq:scalar}
  S(N, M|\bm{x}, \bm{y}) & =
  \bra{\bm{0}} \prod_{j=1}^N \mathbb{C}(x_j) \prod_{j=1}^N \mathbb{B}(y_j) \ket{\bm{0}} \\ 
 & = \frac{ \det H(\bm{x},\bm{y})}{ \prod_{i<j}(x_i - x_j)(y_i - y_j)} \; ,
\end{split}
\end{equation}
where \(H\) is an \(N\times N\) matrix with components
\begin{equation}
\label{eq:h-matrix}
  H_{ij} = H(x_i, y_j) 
  =\frac{1 - (x_i y_j)^{ M + N}}{1 - x_i y_j }\; .
\end{equation}

@@

\subsubsection{Matrix Models}

It is also interesting to notice that the particular case when \(M\to
\infty\), but with a finite number of particles \(N\). The
partitions \(\lambda \in [N, \infty]\) satisfy the condition
\(\ell(\lambda) \leq N\). Therefore, the scalar
product~(\ref{eq:scalar_exp}) becomes
\begin{equation}
  S_N(\bm{t}, \bm{t}') = \sum_{\substack{\lambda \\ \ell(\lambda) \leq N}}
  s_\lambda(\bm{t})  s_\lambda(\bm{t}') \; .
\end{equation}
From~\cite{Zabrodin:2010ii}, we know that this expression can be
written as the following integral
\begin{equation}
  S_N(\bm{t}, \bm{t}') =
  \frac{1}{N!} \prod_{\ell=1}^N \oint_{\Gamma_\ell} \frac{dz_\ell}{2 \pi i z_\ell}
  e^{\xi(\bm{t}, z_\ell) - \xi(\bm{t}', z_{\ell}^{-1})} \Delta(z)\Delta(z^{-1})\; ,
\end{equation}
where
\begin{equation}
\xi(\bm{t}, z) = \sum_{k\geq 0} t_k z^k \qquad
\xi(\bm{t}', 1/z) = \sum_{k\geq 0} t'_{-k} z^{-k} \; . 
\end{equation}
See also~\cite{Kharchev:1991gd} for a detailed proof of this relation,
and~\cite{Orlov:2005} for other details.

What is very interesting about this result is what it implies when we 
impose the Bethe equations~(\ref{eq:bethe_eq}). Suppose that 
\(x_j = e^{-ip_j} \in \mathbb{S}^1\), and we set \(\bm{t}' = - \bm{t}^\star\).
In this particular case, we have
\begin{equation}
\xi(\bm{t}, z) - \xi(\bm{t}', 1/z)  = 2 \ \textrm{Re}\left(\sum_k t_k z^k\right)\; .
\end{equation}
Then, the phase model is equivalent to an ensemble of \(N\) 2D Coulomb
particles on a circle. In this case, we have that the quantities
\(\lambda_{\ell}\) are eigenvalues of a matrix \(U\).

Moreover, Zabrodin~\cite{Zabrodin:2010ii}, see citations therein, we also know
that under the rescaling \(t_k \to T_k/ \hbar\), \(t'_k \to T_{-k}/ \hbar\)
and \(N = T_0/ \hbar\), we get the dispersion limit tau function 
\begin{equation}
  F_0(\bm{T}) = \log S_{T_0}(\bm{T}, \bm{T}') + \mathcal{O}(\hbar)\; ,
\end{equation}
a free energy, from the viewpoit of the matrix integral partition
function.  It is not clear how one can use this fact to determine
properties of the integrable model, but it might be possible to study
the analytic structure of the free energy \(F_0\) to gain some
understanding on the Bether roots \(\bm{x}\).  This problem is
currently under further investigation, and we hope to report some new
results elsewhere.

\subsubsection{KP tau function}
Finally, one may also observe that if we fix one set of coordinates,
say \(\bm{y}\), then we write
\begin{equation}
 S(N,M|\bm{t}) 
 = \sum_{\lambda} s_\lambda(\bm{y})  s_\lambda(\bm{t}) 
 \equiv \sum_{\lambda} c_\lambda(\bm{y})  s_\lambda(\bm{t}) \; ,
\end{equation}
with coefficients \(c_\lambda(\bm{y}) = \det (h_{\lambda_i-i +j}(\bm{y}))\).
But now, it is a trivial to notice that these are Plücker
coordinates in the Jacobi-Trudi form, see~\cite{Miwa2000, Alexandrov:2012tr}.

@@

\subsection{Correlation functions}
In~\cite{Bogoliubov2005}, Bogoliubov has also shown that the
correlation functions
\begin{subequations}
\begin{equation}
\begin{split}
\label{eq:correlation}
  P_m(N, M|\bm{x}, \bm{y}\setminus \{y_N\})
  & = \bra{0} \prod_{j=1}^N \mathbb{C}(x_j)
  \prod_{k=1}^{N-1} \mathbb{B}(y_k) \phi_m^\dagger \ket{0}\\
  & =  \prod_{j=1}^N x_{j}^{M/2} \prod_{k=1}^{N-1} y_{j}^{M/2}
  \bra{0} \prod_{j=1}^N C(1/x_j) \prod_{k=1}^{N-1} B(y_k) \phi_m^\dagger \ket{0}
\end{split}
\end{equation}
can be written as
\begin{equation}
  P_m(N, M|\bm{x}, \bm{y}\setminus \{y_N\}) = 
  \frac{(-1)^{N-1}}{y_N^{(N-1)/2}} \prod_{j=1}^N x_{j}^{M/2}
  \prod_{k=1}^{N-1} y_{j}^{M/2}
  \left( \prod_{t<N} \frac{y_N - y_t}{y_t} \right) \frac{\det Q}{\det H} S(N,M|\bm{x}, \bm{y})
\end{equation}
\end{subequations}
where the \(N\times N\) matrix \(Q\) has components 
\begin{equation}
 Q_{jN} = x_j^{(M + N - 1- 2m)/2} \quad  \textrm{and} \quad 
 Q_{jk} = H_{jk} \; , 
\end{equation}
where \(H_{jk} = H(x_j, y_k)\) are the components of the matrix \(H\) in~(\ref{eq:h-matrix}). 
Since the components \(Q_{jN}\) are independent of the coordinates \(y\), 
we cannot write the above expression as a Toda hierarchy tau
function.

We already know from~(\ref{eq:scalar}) that 
\begin{equation}
   S(N,M|\bm{x}, \bm{y}) =  \Delta(x) \Delta(y) \det H \; ,
\end{equation}
therefore we have
\begin{equation}
  P_m(N, M|\bm{x}, \bm{y}\setminus{y_N})  \propto
  \frac{\det Q}{\Delta(x)} \equiv \mathcal{P}_m(N,M|\bm{x}, \bm{y})\; ,
\end{equation}
and we are going to treat the coordinates \(\{ \bm{y} \}\) as
a set of \(N\) fixed parameters. Furthermore, we write \(N\)
functions \(\bm{F}(z) = (F_1, \dots, F_N)\) defined as
\begin{equation}
  F_j (z) = H(z, y_j) \quad \textrm{if} \ j \neq N\; , \qquad \textrm{and}\qquad 
   F_N (z)  = z^{(M + N - 1 - 2m)/2} \; .
\end{equation}
As before, we expand \(F_j(z)\) as the geometric sum
\begin{equation}
  F_j(z) = \frac{1 - (y_j z)^{M + N}}{1 - y_j z} = \sum_{n=0}^{N + M - 1} (y_j z)^n
  \equiv \sum_{n=0}^{M + N -1} f_{j, n} z^n\; , 
\end{equation}
where we will also assume the \(M + N - 1 \in 2 \mathbb{Z}\).

With these definitions, we conclude that
\begin{equation}
\mathcal{P}_m(N,M|\bm{x}, \bm{y}) =\frac{\det_{jk} F_j(x_k)}{\Delta(x)}\; .
\end{equation}
From the expansion 
\begin{equation}
  \det_{jk} F_j(x_k) = \det_{jk} \left(  \sum_{n=0}^{M + N -1} f_{j, n} x_k^n \right) \; .
\end{equation}
and with the aid of the Cauchy-Binet formula, we write  
\begin{equation}
    \mathcal{P}_m(N,M|\bm{x}, \bm{y})
  = \sum_{0\leq \ell_N\leq \dots \leq \ell_1 \leq N+M}
  \frac{\det_{jk}(f_{j, \ell_k}) \det_{jk}(x_k^{\ell_j})}{\Delta(x)}\; . 
\end{equation}
Now we basically use the definition of the Schur polynomials and the
Jacobi-Trudi expression of Plücker coordinates, see~\cite{Alexandrov:2012tr}. 
Finally, this last expression becomes
\begin{equation}
\mathcal{P}_m(N,M|\bm{x}, \bm{y}) =
\sum_{\lambda} c_\lambda(\bm{y}) s_\lambda(\bm{x}) \; ,
\end{equation}
where \(c_\lambda \equiv \det_{jk}(y_j^{\ell_k})\),  \(\ell_k = \lambda_k - k +N\). Therefore,
we have that this expression is a KP tau function.

From the coordinate expansion of the Bethe
vectors~\cite{Bogoliubov2005, Tsilevich:2006}, it is easy to show that
\begin{equation}
\prod_{j=1}^N \mathbb{B}(x_j)\ket{\lambda}  = \sum_{\mu \supset \lambda } s_{\mu/\lambda}(\bm{x})\ket{\mu}\qquad 
 \bra{\lambda} \prod_{j=1}^N \mathbb{C}(x_j) = \sum_{\mu \supset \lambda } s_{\mu/\lambda}(\bm{x})\bra{\mu}
\end{equation}
where \(s_{\mu/\lambda}\) are Schur polynomials~\cite{Macdonald:1998}.
Therefore, we can write the correlation function~(\ref{eq:correlation}) as 
\begin{equation}
\begin{split}
  P_m(N, M|\bm{x}, \bm{y}\setminus \{y_N\})
  & = \sum_\lambda \sum_{\mu \supset (m)} s_{\lambda}(\bm{x}) s_{\mu/(m)}(\bm{y}\setminus\{y_N\})\\
  & = \sum_\mu s_{\mu/(m)}(\bm{y}\setminus\{y_N\}) s_{\mu}(\bm{x})\; .
\end{split}
\end{equation}
where we have used that \(\phi^\dagger \ket{\bm{0}} = \ket{(m)}\), and in the second line
we sum over all partitions, since \(s_{\mu/(m)} = 0\), \(\forall \) \(\mu \not \supset (m)\).
This expression also shows that the expresison \(P_m\) can also be written as a tau function.

\subsubsection{Schur functions expansion for other correlation functions}
More generally, let us consider the correlation functions 
\begin{equation}
\begin{split}
  P_{\lambda_1 \lambda_2}(N', N, M|\bm{x}, \bm{y}) & =
  \bra{\lambda_1} \prod_{j=1}^N \mathbb{C}(x_j)
  \prod_{k=1}^{N'} \mathbb{B}(y_k)\ket{\lambda_2} \\
  & = \sum_{\mu} s_{\mu/\lambda_1}(\bm{x}) s_{\mu/\lambda_2}(\bm{y}) \; .
\end{split}
\end{equation}
where \((N' + \ell(\lambda_2) + n_0 = N' + \ell(\lambda_2) + n'_0)\) and 
we have also used that the expression for any Young diagram
\(\mu\) that does not contain \(\lambda_1\) and \(\lambda_2\)
vanishes. Using the elementary property of skew Schur
polynomials~\cite{Macdonald:1998}
\begin{equation}
  \sum_{\mu} s_{\mu/\lambda_1}(\bm{x}) s_{\mu/\lambda_2}(\bm{y}) = \prod_{i,j}\frac{1}{1 - x_i y_j}
 \sum_\nu s_{\lambda_1/\nu}(\bm{x}) s_{\lambda_2/\nu}(\bm{y})\; ,
\end{equation}
and the Cauchy's identity, we have 
\begin{equation}
\begin{split}
  P_{\lambda_1 \lambda_2}(N', N, M|\bm{x}, \bm{y}) & =
 \sum_\mu s_{\mu}(\bm{x}) s_{\mu}(\bm{y})
 \sum_\nu s_{\lambda_1/\nu}(\bm{x}) s_{\lambda_2/\nu}(\bm{y})\; .
\end{split}
\end{equation}
Therefore, this correlation function is nothing but the off-shell
norm~(\ref{eq:scalar}) times a finite sum over skew Schur functions.
This expression shows that the existence of these tau functions is not
a trivial fact of this model.

In fact, it is worth considering other quantities that are not
necessarily tau functions. Consider the state
\begin{equation}
  \ket{\mathcal{Y}} =
 \sum_{\vec{n}} \prod_{j=0}^M \ket{n_j}\quad \sum_j n_j = N\; .
\end{equation}
We have~\cite{Bogoliubov2005} the correlation function 
\begin{equation}
  \prod_{j=1}^N x^{M/2} \bra{\mathcal{Y}} \prod_{j=1}^N B(\bm{x}) \ket{\bm{0}} = 
  \sum_{\lambda\subseteq (N,N,M)} s_\lambda(\bm{x})\; . 
\end{equation}
This sum of Schur polynomials are not tau functions of the KP hierarchy.
